---
title: "Lasting Special Places (LSP) – Sense of Place (SP) Subgoal Data Prep"
output:
  github_document:
    toc: true
    toc_depth: 3
params: 
    datasource: csv
always_allow_html: true
---


```{r lsp preamble, echo = FALSE, include = FALSE, error = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE, results = "show", fig.width = 9.5, fig.height = 6)
source(here::here("R", "setup.R"))
source(here::here("R", "spatial.R"))

dir <- here::here("prep", "LSP", "v2021")
bkgd_path <- here::here("supplement", "goal_summaries", "LSP.Rmd")
glossary_path <- here::here("supplement", "glossary.Rmd")
```

<br>

## 1. Background

```{r glossary, child = glossary_path, results = "asis", echo = FALSE}
```

<br>

```{r lsp background, child = bkgd_path, results = "asis", echo = FALSE}
```

<br/>

## 2. Data

This prep document is used to generate and explore the following data layers:

- `LAYER1_bhi2019.csv` 
- `LAYER2_bhi2019.csv` 
- `LAYER3_bhi2019.csv` 

These are saved to the `layers` folder. Saved to `prep/LSP/v2021/intermediate` are intermediate datasets: `DATASETINTEMED1.csv` and `DATASETINTEMED2.csv`. All these are derived from or informed by the raw datasets from RAW DATA SOURCES.

<br>

### 2.1 Datasets with Sources
<br/>

#### 2.1.1 DATA NAME/VARIABLE {-}

**DATASET SUBSET/SUBGROUP NAME/VARIABLE**  
<!-- dataset save location BHI_share/BHI 3.0/Goals/ -->

```{r echo = FALSE}
tab <- t(data.frame(
  c("[option]:", "[specification]"), 
  c("[option]:", "[specification]")))

colnames(tab) <- c("Option", "Specification")
rownames(tab) <- NULL

knitr::kable(tab, caption = "Source: [ABC database](URL/LINK) <br/> Downloaded DATE by NAME")
```

---

<br/>


### 2.2 Centralization & Normalization

```{r lsp load raw data, echo = TRUE, message = FALSE, warning = FALSE, results = "hide"}
## root location of the raw data
dir_rawdata <- file.path(dir_B, "Goals", "LSP")
```

#### 2.2.1 Rename Fields/Variables

#### 2.2.2 Standardize Units

#### 2.2.3 Save to BHI Database

<br/>

### 2.3 Initial Data Exploration

#### 2.3.1 Compare versus Previous Years Data

#### 2.3.2 Timeseries Plots

```{r CODE CHUNK WITH FIGURE OR GRAPH, message = FALSE, echo = TRUE, fig.width = 9.5, fig.height = 4.5}
```

#### 2.3.3 Map


<br/>

## 3. Prep: Wrangling & Derivations, Checks/Evaluation, Gapfilling

<br/>

## 4. Visualizing Data Layers

### 4.1 Map

```{r SOME MAP, message = FALSE, echo = TRUE, fig.width = 9.5}

```

<br>

## 5. Considerations for `BHI4`

<br>

## 6. References


